String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-02 2023-1-19/sections/algorithms.tex
algorithm

The framework is shown in Algorithm \ref{alg:emptiness}, we simplify the input CEFAs at line 1 because they may have too many transitions and registers to solve. The main idea of the simplification is treating the alphabet as unary and the symbolic updates as characters of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 3 and line 4. An off-the-shelf SMT solver gives the result of LIA $\varphi_{under}\wedge \psi$. If the LIA is satisfiable, we return $true$ at line 5 because the CEFAs under the LIA $\psi$ are not empty. Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly. \newline

@@

    \For{$bound \gets 1, MaxBound$ }
    \State $\varphi_{under}\gets \algfun{underApprox}(simpliAuts, bound)$
    \If{$\varphi_{under}\wedge \psi$ is sat}

@@

    \State $R=(\cdots r_{j-1}r_{j+1}\cdots)$ \Comment remove $r_j$ from the register vector $R$

@@

    \State $\myvec{v'}=(\cdots\myvec{v'}[j-1]\myvec{v'}[j+1]\cdots)$ \Comment remove $\myvec{v'}[j]$ from each vector $\myvec{v'}$

@@

    \State $\theta = \theta\wedge r_i = r_j$

@@

\subsubsection{Under-Approximation} \label{subsec:under-approx}
\denghang{revising} \newline
The purpose of under-approximation is to accelerate the solving process of satisfiable instances. With the observation that the accepted string words in practical instances are usual short, we under-approximate the \emptinessprob{} by adding length bounds on string variables. We use string length as the bound and probe the CEFA from low to high bound. This section discusses the situation in which the string length bound is fixed. The main idea is to enumerate all possible runs ending in accepting states with lengths less than the fixed bound. Although the number of possible runs is an exponential growth of the bound, we usually solve practical satisfiable instances within 1s. As shown in the Algorithm \ref{alg:underApprox}, for each CEFA, we enumerate all runs whose length is less than the bound and compute the value of registers by the sum of vectors on the run at line 4. If the run is ended with an accepting state, then we compute the updates of registers by the sum of vectors on the run at line 6. Together with the accepting condition of the accepting state, we add the updates of registers to the under-approximation formula at line 7.
\begin{algorithm}[h]
  \caption{$\algfun{underApprox}(auts, bound)$}
  \label{alg:underApprox}
  \begin{algorithmic}[1]
    \Require The CEFAs $auts$ and string length $bound$
    \Ensure The linear integer arithmetic $\varphi_{under}$ representing under-approximation of the register values of $auts$
    \Statex
    \State $\varphi_{under} \gets false$
    \For{$\aut \in auts$}
    \State Suppose $\aut = (Q, \Sigma, \delta, q_I, F, (r_1, \cdots, r_m), \theta)$
    \ForAll{run $q_0q_1\cdots q_{n}$ whose length is less than $bound$}
    \If{$q_{n}\in F$}
    \State Let $\myvec{V}\leftarrow$ the sum of the vectors on the run
    \State $\varphi_{under} \gets \varphi_{under} \vee(\theta(q_n)\bigwedge\limits_{i\in[1,m]} r_i = \myvec{V}[i])$
    \EndIf
    \EndFor
    \EndFor
    \State \textbf{return} $\varphi_{under}$
  \end{algorithmic}
\end{algorithm}
\begin{example}
  Consider the automaton in Fig. \ref{subfig:aut_x} again. The solution of the Parikh image of it leads to a complete but inefficient result since the quantifier-free linear integer arithmetic formulas are solved in exponential time \cite{parikh_compute}. Under-approximation can accelerate the solving process. When we set the bound to 2, the under-approximation program will enumerate all runs whose length is 2. One of them is $q_I\xrightarrow[(1,1)]{a} q_1\xrightarrow[(0,1)]{b} q_f$. The corresponding registers' values are $r_1 = 1+0 = 1, r_2 = 1+1 = 2$. Checking for $r_1= 1\wedge r_2=2\wedge r_2=i\wedge 1\leq r_1\leq100$ is a lightweight process because the values of $r_1$ and $r_2$ are fixed.
\end{example}

>>>>

The framework is shown in Algorithm \ref{alg:emptiness}, we simplify the input CEFAs at line 1 because they may have too many transitions and registers to solve. The main idea of the simplification is treating the alphabet as unary and the symbolic updates as characters of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 2 and line 3. The under-approximation algorithm explore satisfiable register values from low string-length to high string-length. If it finds a solution \denghang{todo} Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly. \newline

@@

    % \For{$bound \gets 1, MaxBound$ }
    % \State $\varphi_{under}\gets \algfun{underApprox}(simpliAuts, bound)$
    % \If{$\varphi_{under}\wedge \psi$ is sat}
    % \State \textbf{return} $true$
    % \EndIf
    % \EndFor
    \If{\algfun{underApproxSolve}($simpliAuts, \psi, MaxBound$) is true}

@@

    \State $R\gets(\cdots r_{j-1}r_{j+1}\cdots)$ \Comment remove $r_j$ from the register vector $R$

@@

    \State $\myvec{v'}\gets(\cdots\myvec{v'}[j-1]\myvec{v'}[j+1]\cdots)$ \Comment remove $\myvec{v'}[j]$ from each vector $\myvec{v'}$

@@

    \State $\theta \gets\theta\wedge r_i = r_j$

@@

\subsubsection{Under-Approximation} \label{subsec:under-approx}
\denghang{revising} \newline
The purpose of under-approximation is to accelerate the solving process of satisfiable instances. With the observation that the accepted string words in practical instances are usual short, we under-approximate the \emptinessprob{} by adding length bounds on string variables. We use string length as the bound and probe the CEFA from low to high bound. This section discusses the situation in which the string length bound is fixed. The main idea is to enumerate all possible runs ending in accepting states with lengths less than the fixed bound. Although the number of possible runs is an exponential growth of the bound, we usually solve practical satisfiable instances within 1s. As shown in the Algorithm \ref{alg:underApprox}, for each CEFA, we enumerate all runs whose length is less than the bound and compute the value of registers by the sum of vectors on the run at line 4. If the run is ended with an accepting state, then we compute the updates of registers by the sum of vectors on the run at line 6. Together with the accepting condition of the accepting state, we add the updates of registers to the under-approximation formula at line 7.
  \caption{$\algfun{underApproxSolve}(auts, bound, \psi)$}
  \label{alg:underApprox}
    \Require The CEFAs $auts$, max string length $bound$, other LIA formula $\psi$
    % \Ensure The linear integer arithmetic $\varphi_{under}$ representing under-approximation of the register values of $auts$
    \Ensure \emph{true} if finding a solution, \emph{false} otherwise
    \State $\varphi_{under}$ is an incremental LIA formula as under approximation, initialized to $\psi$
    \State $\varphi_\aut$ is an incremental LIA formula for each automaton $\aut$, initialized to $false$
    \State $\theta$ is the conjunction of all accepting conditions of $auts$
    \State $lastStates(\aut)$ is the set of (last state, vector's sum) pairs of $\aut$, initialized to $\{(q_I, 0_n)\}$ 
    \While {$bound \geq 0$}
        \For{$(q, \myvec{V}) \gets lastStates(\aut)$}
          \If{$q\in F$}
            \State $\varphi_{\aut} \gets \varphi_{\aut} \bigvee\limits_{i\in[1,m]} r_i = \myvec{V}[i])$
          \State Remove $(q, \myvec{V})$ from $lastStates(\aut)$
          \State Add $(q', \myvec{V}+\myvec{v})$ to $lastStates(\aut)$ for each transition $q\xrightarrow[\myvec{v}]{a} q'\in \aut$
        \State $\varphi_{under}\gets \varphi_{under}\wedge\varphi_\aut$
      \If{$\varphi_{under}\wedge \theta$ is satisfiable}
        \State \textbf{return} \emph{true}
      \State $bound \gets bound - 1$
    \EndWhile
    \State \textbf{return} \emph{false}
    % \For{$\aut \in auts$}
    % \State $\varphi_\aut\gets false$ 
    % \State Suppose $\aut = (Q, \Sigma, \delta, q_I, F, (r_1, \cdots, r_m), \theta)$
    % \ForAll{run $q_0q_1\cdots q_{n}$ whose length is less than $bound$}
    % \If{$q_{n}\in F$}
    % \State Let $\myvec{V}$ to be the sum of the vectors on the run
    % \State $\varphi_{\aut} \gets \varphi_{\aut} \vee(\theta(q_n)\bigwedge\limits_{i\in[1,m]} r_i = \myvec{V}[i])$
    % \EndIf
    % \EndFor
    % \State $\varphi_{under} \gets \varphi_{under}\wedge \varphi_{\aut}$
    % \EndFor
    % \State \textbf{return} $\varphi_{under}$
  Consider the automaton in Fig. \ref{subfig:aut_x} again. The solution of the Parikh image of it leads to a complete but inefficient result since the quantifier-free linear integer arithmetic formulas are solved in exponential time \cite{parikh_compute}. Under-approximation can accelerate the solving process. When we set the bound to 2, the under-approximation program will enumerate all runs whose length is 2. One of them is $q_I\xrightarrow[(1,1)]{a} q_1\xrightarrow[(0,1)]{b} q_f$. The corresponding registers' values are $r_1 = 1+0 = 1, r_2 = 1+1 = 2$. Checking for $r_1= 1\wedge r_2=2\wedge r_2=i\wedge 1\leq r_1\leq100$ is a lightweight process because the values of $r_1$ and $r_2$ are fixed.
